---
title: "Adenomyosis spatial transcriptomics"
subtitle: "Reviewers comments - notch signaling and ciliogenesis, hypoxia"
author: "Emily Johnson"
date: today
date-format: short
format:
  html:
    self-contained: true
    theme: litera
    toc: true
editor: source
code-block-bg: true
code-block-border-left: "#31BAE9"
---

# Notch signaling and ciliogenesis analysis

## Load libraries

```{r}
#| output: false
#| message: false
library(tidyverse)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggplot2)
library(cowplot)
library(pheatmap)
library(RColorBrewer)
library(openxlsx)
source("src/utilityFunctions.R")
```

## Load data

```{r}

# Load differential expression results
limma_GeoDiff_results <- readRDS("results/differential_expression/limma_GeoDiff_pairwise_comparisons.RDS")

# Load functional enrichment results
GeoDiff_limma_enrichment <- readRDS("results/functional_enrichment/GeoDiff_limma_enrichment.rds")

# Load cilium core enrichment genes
cilium_genes <- read.csv("results/functional_enrichment/cilium_core_enrichment_genes.csv")

```

## Extract Notch pathways from enrichment results

```{r}

# Check if Notch signaling pathway is enriched in GSEA results
panck_enrichment <- GeoDiff_limma_enrichment$PanCK

CD45_enrichment <- GeoDiff_limma_enrichment$CD45

CD31_enrichment <- GeoDiff_limma_enrichment$CD31

stroma_enrichment <- GeoDiff_limma_enrichment$Stroma

# Extract Notch-related pathways
notch_pathways <- panck_enrichment@compareClusterResult %>%
  filter(grepl("notch|NOTCH", Description, ignore.case = TRUE))

notch_pathways_CD45 <- CD45_enrichment@compareClusterResult %>%
  filter(grepl("notch|NOTCH", Description, ignore.case = TRUE))

notch_pathways_CD31 <- CD31_enrichment@compareClusterResult %>%
  filter(grepl("notch|NOTCH", Description, ignore.case = TRUE))

notch_pathways_stroma <- stroma_enrichment@compareClusterResult %>%
  filter(grepl("notch|NOTCH", Description, ignore.case = TRUE))


write.csv(notch_pathways, "results/functional_enrichment/notch_reactome_pathways_panCK.csv")
write.csv(notch_pathways_CD45, "results/functional_enrichment/notch_reactome_pathways_CD45.csv")
write.csv(notch_pathways_CD31, "results/functional_enrichment/notch_reactome_pathways_CD31.csv")
write.csv(notch_pathways_stroma, "results/functional_enrichment/notch_reactome_pathways_stroma.csv")


if (nrow(notch_pathways) > 0) {
  print(notch_pathways %>%
          dplyr::select(Cluster, Description, NES, pvalue, p.adjust, core_enrichment) %>%
          arrange(p.adjust))
  
  # Extract core enrichment genes from Notch pathways
  notch_pathway_genes <- notch_pathways %>%
    dplyr::select(Description, core_enrichment) %>%
    mutate(genes = strsplit(core_enrichment, "/")) %>%
    unnest(genes) %>%
    dplyr::select(Description, genes)
  
  print(head(notch_pathway_genes, 20))
  
  # Get unique genes from pathways
  pathway_notch_genes <- unique(notch_pathway_genes$genes)
  
  (sprintf("\n\nTotal unique genes from Notch pathways: %d\n", 
              length(pathway_notch_genes)))
  
} else {
  print("No Notch pathways significantly enriched\n")
  pathway_notch_genes <- NULL
}

```

## Define Notch signaling genes

Use both curated list and pathway-derived genes.

```{r}

# Notch receptors
notch_receptors <- c("NOTCH1", "NOTCH2", "NOTCH3", "NOTCH4")

# Notch ligands
notch_ligands <- c("DLL1", "DLL3", "DLL4", "JAG1", "JAG2")

# Notch target genes and downstream effectors
notch_targets <- c(
  "HES1", "HES5", "HES7", "HEY1", "HEY2", "HEYL",  # Canonical targets
  "NRARP", "LFNG", "MFNG", "RFNG",                  # Fringe genes
  "MAML1", "MAML2", "MAML3",                        # Co-activators
  "RBPJ", "RBPJL",                                  # DNA binding proteins
  "DTX1", "DTX2", "DTX3", "DTX4"                    # Deltex proteins
)

# Combine curated genes with pathway-derived genes
if (!is.null(pathway_notch_genes)) {
  notch_genes <- unique(c(notch_receptors, notch_ligands, notch_targets, pathway_notch_genes))
  print(sprintf("Total Notch pathway genes (curated + pathway): %d\n", 
              length(notch_genes)))
} else {
  notch_genes <- c(notch_receptors, notch_ligands, notch_targets)
  print(sprintf("Total Notch pathway genes (curated only): %d\n", 
              length(notch_genes)))
}

```

## Extract key Notch genes from differential expression

Identify which Notch genes are actually differentially expressed.

```{r}

# Focus on PanCK epithelial compartment (where ciliation is relevant)
panck_contrasts <- limma_GeoDiff_results[grep("PanCK", names(limma_GeoDiff_results))]

# Extract Notch gene expression
notch_expr <- panck_contrasts %>%
  map_df(~{
    res <- .x[row.names(.x) %in% notch_genes, , drop = FALSE]
    if(nrow(res) > 0) {
      res %>%
        rownames_to_column("Gene") %>%
        dplyr::select(Gene, logFC, adj.P.Val, t)
    } else {
      NULL
    }
  }, .id = "Contrast")

# Add significance flag
notch_expr <- notch_expr %>%
  mutate(
    Significant = adj.P.Val < 0.05 & abs(logFC) > 0.5,
    Direction = case_when(
      Significant & logFC > 0 ~ "Up",
      Significant & logFC < 0 ~ "Down",
      TRUE ~ "Not significant"
    )
  )

# Get significantly DE Notch genes for correlation analysis
sig_notch_genes <- notch_expr %>%
  filter(Significant) %>%
  pull(Gene) %>%
  unique()


```

```{r}

# Summary by gene printegory
notch_summary <- notch_expr %>%
  filter(Significant) %>%
  mutate(
    printegory = case_when(
      Gene %in% notch_receptors ~ "Receptor",
      Gene %in% notch_ligands ~ "Ligand",
      Gene %in% notch_targets ~ "Target/Effector"
    )
  ) %>%
  group_by(Contrast, printegory, Direction) %>%
  summarise(n = n(), .groups = "drop")

print("\nSummary of dysregulated Notch genes by printegory:\n")
print(notch_summary)

```

## Visualise Notch gene expression

```{r}

# Create heatmap of Notch gene expression
notch_matrix <- notch_expr %>%
  dplyr::select(Contrast, Gene, logFC) %>%
  pivot_wider(names_from = Gene, values_from = logFC, values_fill = 0) %>%
  column_to_rownames("Contrast") %>%
  as.matrix()

# Only include genes with at least one significant change
sig_notch_genes <- notch_expr %>%
  filter(Significant) %>%
  pull(Gene) %>%
  unique()

if (length(sig_notch_genes) > 0) {
  notch_matrix_filtered <- notch_matrix[, sig_notch_genes, drop = FALSE]
  
  notch_heatmap <- pheatmap(
    notch_matrix_filtered,
    color = colorRampPalette(c("#6D9EC1", "white", "firebrick3"))(100),
    breaks = seq(-2, 2, length.out = 101),
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    main = "Notch Pathway Gene Expression\n(Adenomyosis vs Endometrium)",
    fontsize = 10,
    cellwidth = 15,
    cellheight = 15
  )
  
  print(notch_heatmap)
}

```

```{r}

# Bar plot showing direction of change
notch_barplot <- ggplot(notch_expr %>% filter(Significant),
                        aes(x = Gene, y = logFC, fill = Direction)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Contrast, scales = "free_x", ncol = 2) +
  scale_fill_manual(values = c("Up" = "firebrick3", "Down" = "#6D9EC1")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Differential Expression of Notch Pathway Genes",
       subtitle = "Adenomyosis Lesion Epithelium vs Endometrium",
       x = "Gene", y = "log2 Fold Change")

print(notch_barplot)

```

## Correlation between Notch genes and cilium genes

```{r}

# Load expression matrix
target_spatial_data_geoDiff <- readRDS("results/GeoDiff_normalised_spatial_data.RDS")
expr_mat <- target_spatial_data_geoDiff@assayData$normmat
pheno <- pData(target_spatial_data_geoDiff)

# Filter for PanCK epithelial samples only
panck_samples <- pheno %>%
  filter(segment == "PanCK") %>%
  rownames_to_column("sampleID") %>%
  pull(sampleID)

expr_panck <- expr_mat[, panck_samples]

```

```{r}

# Get unique cilium genes
unique_cilium_genes <- unique(cilium_genes$genes)

# Use significantly DE Notch genes if available, otherwise use all
notch_genes_for_cor <- if(length(sig_notch_genes) > 0) sig_notch_genes else notch_genes

# Calculate correlations between Notch genes and cilium genes
# Only for genes present in both lists
notch_present <- notch_genes_for_cor[notch_genes_for_cor %in% rownames(expr_panck)]
cilium_present <- unique_cilium_genes[unique_cilium_genes %in% rownames(expr_panck)]

print(sprintf("Notch genes present in data: %d/%d\n", 
            length(notch_present), length(notch_genes_for_cor)))
print(sprintf("Cilium genes present in data: %d/%d\n",
            length(cilium_present), length(unique_cilium_genes)))

```

```{r}

# Calculate correlation matrix
if (length(notch_present) > 0 & length(cilium_present) > 0) {
  
  notch_cilium_cor <- cor(
    t(expr_panck[notch_present, , drop = FALSE]),
    t(expr_panck[cilium_present, , drop = FALSE]),
    use = "pairwise.complete.obs"
  )
  
  # Find strongest correlations
  cor_df <- notch_cilium_cor %>%
    as.data.frame() %>%
    rownames_to_column("Notch_gene") %>%
    pivot_longer(-Notch_gene, names_to = "Cilium_gene", values_to = "Correlation") %>%
    filter(abs(Correlation) > 0.5) %>%
    arrange(desc(abs(Correlation)))
  
  print("\nStrongest correlations between Notch and cilium genes (|r| > 0.5):\n")
  print(head(cor_df, 20))
  
  # Save full correlation results
  write.csv(cor_df,
            "results/functional_enrichment/notch_cilium_correlations.csv",
            row.names = FALSE)
}

```

### Visualise correlations

```{r}

# Improved correlation heatmap
if (exists("notch_cilium_cor") && length(notch_present) > 0 && length(cilium_present) > 0) {
  
  # Visualise full correlation matrix using ggcorrplot style
  library(ggcorrplot)
  
  # For Visualisation, limit to top correlated cilium genes
  # Calculate mean absolute correlation per cilium gene
  mean_cor <- colMeans(abs(notch_cilium_cor))
  top_cilium <- names(sort(mean_cor, decreasing = TRUE))[1:min(20, length(mean_cor))]
  
  # Subset correlation matrix
  cor_subset <- notch_cilium_cor[, top_cilium, drop = FALSE]
  
  # Create correlation plot
  cor_plot <- ggcorrplot(
    t(cor_subset),  # Transpose so cilium genes on y-axis
    type = "full",
    ggtheme = theme_cowplot(),
    lab = FALSE,  # Don't show correlation values (too many)
    colors = c("#6D9EC1", "white", "firebrick3"),
    legend.title = "Pearson\nCorrelation"
  ) +
    labs(
      title = "Notch-Cilium Gene Expression Correlations",
      subtitle = sprintf("Top %d cilium genes by mean correlation (PanCK epithelium)", 
                        length(top_cilium)),
      x = "Notch Pathway Genes",
      y = "Cilium Genes"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
      axis.text.y = element_text(size = 10)
    )
  
  print(cor_plot)
  
  ggsave(
    filename = "figures/functional_enrichment/notch_cilium_correlation_heatmap.pdf",
    plot = cor_plot,
    width = 10,
    height = 8,
    units = "in",
    dpi = 300,
    bg = "white"
  )
}

```

```{r}

# Alternative: tile plot showing only strong correlations
if (exists("cor_df") && nrow(cor_df) > 0) {
  
  top_correlations_plot <- ggplot(
    head(cor_df, 30), 
    aes(x = reorder(Notch_gene, Correlation),
        y = reorder(Cilium_gene, Correlation),
        fill = Correlation)
  ) +
    geom_tile(color = "white", size = 0.5) +
    scale_fill_gradient2(
      low = "#6D9EC1", 
      mid = "white", 
      high = "firebrick3",
      midpoint = 0, 
      limits = c(-1, 1),
      name = "Pearson\nCorrelation"
    ) +
    theme_cowplot() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
      axis.text.y = element_text(size = 10),
      axis.title.x = element_text(size = 12),
      axis.title.y = element_text(size = 12)
    ) +
    labs(
      title = "Strong Notch-Cilium Gene Correlations",
      subtitle = "Top 30 correlations (|r| > 0.5)",
      x = "Notch Pathway Gene", 
      y = "Cilium Gene"
    )
  
  print(top_correlations_plot)
  
  ggsave(
    filename = "figures/functional_enrichment/notch_cilium_top_correlations.pdf",
    plot = top_correlations_plot,
    width = 8,
    height = 10,
    units = "in",
    dpi = 300,
    bg = "white"
  )
}

```

```{r}

# Summary correlation matrix for key genes only
key_notch <- c("NOTCH1", "NOTCH2", "NOTCH3", "HES1", "HEY1")
key_cilium <- c("RSPH1", "RILPL2", "TEKT2", "FOXJ1", "RFX3")

# Filter for genes present in data
key_notch_present <- key_notch[key_notch %in% rownames(expr_panck)]
key_cilium_present <- key_cilium[key_cilium %in% rownames(expr_panck)]

if (length(key_notch_present) > 0 & length(key_cilium_present) > 0) {
  
  # Calculate correlation between key genes
  key_cor <- cor(
    t(expr_panck[key_notch_present, , drop = FALSE]),
    t(expr_panck[key_cilium_present, , drop = FALSE]),
    use = "pairwise.complete.obs"
  )
  
  # Create publiprintion-ready correlation plot
  key_cor_plot <- ggcorrplot(
    t(key_cor),
    type = "full",
    ggtheme = theme_cowplot(),
    lab = TRUE,
    lab_size = 3.5,
    colors = c("#6D9EC1", "white", "firebrick3"),
    legend.title = "Pearson\nCorrelation"
  ) +
    labs(
      title = "Notch-Cilium Gene Correlations",
      subtitle = "Selected markers in PanCK epithelium",
      x = "Notch Pathway Genes",
      y = "Cilium Genes"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 11, face = "bold"),
      axis.text.y = element_text(size = 11, face = "bold")
    )
  
  print(key_cor_plot)
  
  ggsave(
    filename = "figures/functional_enrichment/notch_cilium_key_genes_correlation.pdf",
    plot = key_cor_plot,
    width = 7,
    height = 5,
    units = "in",
    dpi = 300,
    bg = "white"
  )
}

```



## Co-expression network analysis

```{r}

# Examine co-expression of key Notch receptors with cilium-associated genes
# Focus on NOTCH1, NOTCH2, NOTCH3 (most relevant for epithelial ciliogenesis)

key_notch <- c("NOTCH1", "NOTCH2", "NOTCH3")
key_cilium <- c("RSPH1", "RILPL2", "TEKT2", "FOXJ1", "RFX3")  # Key ciliation genes

# Filter for genes present in data
key_notch_present <- key_notch[key_notch %in% rownames(expr_panck)]
key_cilium_present <- key_cilium[key_cilium %in% rownames(expr_panck)]

if (length(key_notch_present) > 0 & length(key_cilium_present) > 0) {
  
  key_genes <- c(key_notch_present, key_cilium_present)
  
  # Get expression of key genes by region
  expr_key <- expr_panck[key_genes, , drop = FALSE] %>%
    t() %>%
    as.data.frame() %>%
    rownames_to_column("sample") %>%
    left_join(
      pheno %>% rownames_to_column("sample") %>% dplyr::select(sample, region),
      by = "sample"
    )
  
  # Visualise expression patterns
  expr_key_long <- expr_key %>%
    pivot_longer(cols = all_of(key_genes),
                 names_to = "Gene",
                 values_to = "Expression") %>%
    mutate(
      Gene_type = ifelse(Gene %in% key_notch_present, "Notch", "Cilium")
    )
  
  coexpr_plot <- ggplot(expr_key_long, 
                        aes(x = region, y = Expression, fill = Gene_type)) +
    geom_boxplot() +
    facet_wrap(~Gene, scales = "free_y", ncol = 4) +
    scale_fill_manual(values = c("Notch" = "#BEC100", "Cilium" = "#60a8a5")) +
    theme_cowplot() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = "Expression of Key Notch and Cilium Genes",
         x = "Region", y = "Normalised Expression", fill = "Gene Type")
  
  print(coexpr_plot)
}

```



## Save results

```{r}

# Save Notch gene expression results
write.csv(notch_expr, 
          "results/functional_enrichment/notch_gene_expression.csv",
          row.names = FALSE)

# Save correlations
if (exists("cor_df")) {
  write.csv(cor_df,
            "results/functional_enrichment/notch_cilium_correlations.csv",
            row.names = FALSE)
}

```

```{r}

# Save plots
if (exists("notch_heatmap")) {
  pdf("figures/functional_enrichment/notch_pathway_heatmap.pdf",
      width = 10, height = 6)
  print(notch_heatmap)
  dev.off()
}

ggsave(
  filename = "figures/functional_enrichment/notch_gene_barplot.pdf",
  plot = notch_barplot,
  width = 12,
  height = 8,
  units = "in",
  dpi = 300,
  bg = "white"
)

if (exists("coexpr_plot")) {
  ggsave(
    filename = "figures/functional_enrichment/notch_cilium_coexpression.pdf",
    plot = coexpr_plot,
    width = 12,
    height = 6,
    units = "in",
    dpi = 300,
    bg = "white"
  )
}

if (exists("notch_cilium_sprintter")) {
  ggsave(
    filename = "figures/functional_enrichment/notch_cilium_correlation.pdf",
    plot = notch_cilium_sprintter,
    width = 6,
    height = 5,
    units = "in",
    dpi = 300,
    bg = "white"
  )
}

```


# Hypoxia

## Load libraries

```{r}
#| output: false
#| message: false
library(tidyverse)
library(openxlsx)

```

## Load data

```{r}

GeoDiff_limma_enrichment <- readRDS("results/functional_enrichment/GeoDiff_limma_enrichment.rds")
GeoDiff_limma_GO_term_enrichment <- readRDS("results/functional_enrichment/GeoDiff_limma_GO_term_enrichment.rds")

```

## Extract hypoxia related pathways/terms

```{r}

# Set hypoxia search regex
hypoxia_regex <- "hypoxia|hypoxic|HIF|oxygen|oxidative stress|response to decreased oxygen"


# Helper function to extract terms
extract_hypoxia_terms <- function(enrichment_obj, regex = hypoxia_regex) {
  enrichment_obj@compareClusterResult %>%
    filter(grepl(regex, Description, ignore.case = TRUE))
}


# Extract Reactome hypoxia terms
reactome_segments <- names(GeoDiff_limma_enrichment)

reactome_hypoxia_list <- lapply(reactome_segments, function(seg) {
  extract_hypoxia_terms(GeoDiff_limma_enrichment[[seg]])
})

names(reactome_hypoxia_list) <- reactome_segments


# Extract GO hypoxia terms
go_segments <- names(GeoDiff_limma_GO_term_enrichment)

go_hypoxia_list <- lapply(go_segments, function(seg) {
  extract_hypoxia_terms(GeoDiff_limma_GO_term_enrichment[[seg]])
})

names(go_hypoxia_list) <- go_segments


# Export results to Excel
reactome_file <- "results/functional_enrichment/hypoxia_reactome_terms.xlsx"
wb_reactome <- createWorkbook()

for (seg in reactome_segments) {
  addWorksheet(wb_reactome, seg)
  writeData(wb_reactome, seg, reactome_hypoxia_list[[seg]])
}

saveWorkbook(wb_reactome, reactome_file, overwrite = TRUE)


go_file <- "results/functional_enrichment/hypoxia_GO_terms.xlsx"
wb_go <- createWorkbook()

for (seg in go_segments) {
  addWorksheet(wb_go, seg)
  writeData(wb_go, seg, go_hypoxia_list[[seg]])
}

saveWorkbook(wb_go, go_file, overwrite = TRUE)


```

::: {.callout-note collapse="true"}
## Session Info

```{r}
sessionInfo()
```
:::
